Dynamic Prediction of Non-Guassian Outcome with fast Generalized Functional Principal Analysis

### 2023-04-06 Write-ups/Manuscript_edit.tex
edit introduction

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{April 04, 2023}

@@

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, typically
referred to as dynamic prediction. Traditionally, this type of data has
been modeled either marginally, such as generalized estimating
equations, or constitutionally on specific subjects, such as mixed
effect models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
\protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990};
\protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
measurement data''} 2003}). For subject-specific models, predictions are
made based on the correlation between repeated measures from the same
subject and covariates that can be either fixed or varying. However,
most of these methods are based on linear model structure, which is very
limited in terms of flexibility of correlation. When sample size is
large or the density of repeated measures is high, they also tend to
cause severe computational burden
(\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}). In addition,
with generalized, non-Gaussian outcomes, these methods lack the ability
to handle out-of-sample prediction because there is usually no
close-form solution for subject-specific random effects

To address these problems, one may turn to functional mixed effect
models instead when measures are dense across the domain of measures.
Such methods accommodate more flexible correlation structure by modeling
subject-specific random effects as a function. Non-parametric smoothing
(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014}) are often
incorporated to speed up the computation, where random effects are
approximated by spline basis functions or eigenfunctions from functional
principal component analysis (FPCA). The introduction of basis functions
also makes out-of-sample prediction more straightforward. Instead of
estimating subject-specific random effects of new observations, we can
simply estimate coefficients/loadings on the basis function used for
smoothing. Existing research on dynamic prediction with functional data
analysis methods has been focusing on continuous/Gaussian outcomes,
usually modelling subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors.

Unfortunately, fewer papers have focused on its extension to
generalized, non-Gaussian outcomes, such as series of binary or count
outcomes. Existing methods also tend to be very computationally
intensive. For example, Chen et al.
(\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
marginal functional models that is compatible to multi-level,
generalized outcomes. Goldsmith et al.
(\protect\hyperlink{ref-goldsmith2015}{2015}) established a model
framework that takes into account the fixed effect of time-invariant
covariates, with parameters estimated with Bayesian method in
\emph{Stan}. Gertheiss et al.
(\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
by directly applying FPCA methods to generalized functions, and proposed
to address this problem using a two-stage joint estimation strategy.
Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
variational algorithm for FPCA of binary and count data. In terms of
implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
proposed a fast, efficient way to fit GFPCA on binary data using EM
algorithm, accompanied by the an open source R package \emph{registr}.

In this paper, we aim to develop a fast, scalable method for dynamic
prediction of generalized functional outcome based on functional mixed
effect model with FPCA smoothing. Specifically, ``Generalized''
indicates that the outcome follows non-Gaussian exponential family
distribution, and the term ``functional'' indicates that repeated
measure are collected densely across the study domain (e.g., time).
Section 2 explains the implementation procedure of the proposed method.
In Section 3, we illustrate the performance and efficiency of our
proposed method in a simulation study. In Section 4, we apply this
method to a real world dataset from a weight loss study. Section 5
presents a discussion of advantages and limitation of the proposed
method.

@@

where g is an appropriate link function, \(\beta_0(t)\) is the
population mean of latent function, and \(b_i(t)\) is a subjects-level
random effect function following a zero-mean Gaussian process.

According to the Karhunen-Loève theorem, the unobserved random effect
function \(b_i(t)\) can be represented as the linear combination of a
infinite set of orthogonal functions:
\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). We approximate this
function by a finite set of \(\phi_{k}(t)\), such that
\(b_i(t)=\sum_{k=1}^{K}\xi_{ik}\phi_{k}(t)\).

In practice, this approximation could be implemented through an FPCA
procedure on \(\eta_i(t)\). In this case
\(\phi_{k}(t),\ k \in \{1,...,K\}\) are the first K eigenfunctions that
explain the most variation in \(\eta_i(t)\). These eigenfunctions, also
referred to as Principal Component (PC) functions, are orthogonal are
can be orthonormal with proper scaling. \(\xi_{ik}\) are
subject-specific scores (or loadings) on each PC function. According to
properties of FPCA method, these scores are mutually independent over
both subject (\(i\)) and eigenfunctions (\(k\)). That is, each
\(\xi_{ik}\) follows normal distribution \(N(0, \lambda_k)\), where
\(\lambda_k\) is the kth eigenvalues.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\). The
  number of observations from the same subject in bin \(s\) is \(n_s\),
  and \(t_{sj}\) refers to the jth observation point in bin \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t_{sj}))] = \beta_0(s)+b_i(s)\),
  \(j\in\{1, ..., n_s\}\). From this series of models we can estimate
  the individual latent functions at every bin:
  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions:
  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
  where \(\epsilon_i(t)\) indicates random noise introduced by the
  binning process in previous steps. This random noise is assumed to
  follow a zero-mean Gaussian process, such that at each specific point
  t, \(\epsilon_i(t) \sim N(0, \sigma^2)\). The FPCA output includes
  estimates of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
  and residual variance \(\hat{\sigma}^2\).
\end{enumerate}

@@

As a reference method, we compare our method to Generalized Linear Mixed
Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of
the fastest existing method developed for dynamic prediction of repeated
generalized outcomes. Just like many mixed models, this method is very
limited in terms of flexibility. For example, the model used for
prediction of our simulated datset would simply be an linear model with
one covariate indicating observation time:
\(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
of this mixed model can be increased using spline functions, the
dimension of spline basis is also restricted by computational ability,
and is unfeasible to implement under the scale of our simulated data or
the complexity of our proposed method.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/Figure-1.pdf}
\caption{Predicted track of four randomly selected subjects from one
simulated dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicated predicted latent
function tracks, and color indicates different observation time.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Manuscript_files/figure-latex/Figure_appl-1.pdf}
\caption{Predicted track of four randomly selected subjects from the
daily weigh-in dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicated predicted latent
function tracks, and color indicates different observation time.}
\end{figure}

@@

The simulation study and case study of daily weigh-in data above have
revealed the proposed method can achieve better predictive performance
with much less computational burden. Compared to existing methods based
on linear mixed model, fGFPCA can accommodate flexible correlation
structure between repeated measure, which is particularly useful when
outcome is generated from latent functions with wiggly or cyclic
pattern. As in the simulation setting in Section 3, the binary outcomes
are generated from a cyclic process. fGFPCA has much lower ISE and
higher AUC than GLMM in this case, and its prediction accuracy also
improves as more outcomes are collected, which is a desirable property
for dynamic prediction where the datasets would be periodically updated
with newly collected data. Moreover, fGFPCA also significantly reduce
time spent on model fitting and predict, which is beneficial when
repeated measures are dense or sample size is large. In fact, as far as
we are aware of, the proposed method is the only feasible method that
can handle the scale of the simulated datasets with compatible
flexibility.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-leroux2016}{}}%
Leroux, A., Xiao, L., Crainiceanu, C., and Checkley, W. (2018),
{``Dynamic prediction in functional concurrent regression with an
application to child growth,''} \emph{Statistics in medicine}, 37,
1376--1388.

@@

@@

>>>>

\title{Dynamic Prediction of Non-Guassian Outcome with fast Generalized
Functional Principal Analysis}
\author{Ying Jin \and Andrew Leroux}
\date{April 06, 2023}

@@

Biomedical investigators are often interested in predicting future
observations of subjects based on their historical data, typically
referred to as dynamic prediction. Traditionally, this type of data has
been modeled either marginally, such as generalized estimating
equations, or conditionally on specific subjects, such as mixed effect
models (\protect\hyperlink{ref-Laird1982}{Laird and Ware 1982};
\protect\hyperlink{ref-liang1986}{LIANG and ZEGER 1986};
\protect\hyperlink{ref-lindstrom1990}{Lindstrom and Bates 1990};
\protect\hyperlink{ref-davidian2003}{{``Nonlinear models for repeated
measurement data''} 2003}). For subject-specific models, predictions are
made based on the correlation between repeated measures from the same
subject and covariates that can be either fixed or varying. Software and
methods for dynamic prediction have been developed for Gaussian
{[}leroux2016; Shang (\protect\hyperlink{ref-shang2017}{2017}); Goldberg
et al. (\protect\hyperlink{ref-goldberg2014}{2014}){]}, non-Gaussian
data (\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-GLMMadaptive}{Rizopoulos 2022}), and
time-to-event outcomes (\protect\hyperlink{ref-jmbook}{Rizopoulos 2012};
\protect\hyperlink{ref-suresh2017}{Suresh et al. 2017}). However, with
the notable exception of Leroux et al.
(\protect\hyperlink{ref-leroux2016}{2018a}), existing methods are either
limited in their flexibility (both fixed and random effect structure,
e.g.~random slope and intercept), computationally unfeasible for large
datasets, or both. These issues are particularly acute for data which
are measured densely, as is often the case with functional data, or when
data are non-Gaussian.

While methods with accompanying software for modeling complex functional
data with non-linear fixed and random effects exist for both Gaussian
and non-Gaussian data, software is largely limited to the refund package
in the statistical software R
(\protect\hyperlink{ref-Scheipl2014}{Scheipl et al. 2014};
\protect\hyperlink{ref-wood2014}{Wood and Scheipl 2014};
\protect\hyperlink{ref-refundpkg}{\textbf{refundpkg?}}). However, these
methods and software implementations do not currently allow for dynamic
prediction. In addition to the methods implemented in the refund
package, other work has been done on the fitting of functional
regression models to non-Gaussian functional data. For example, Chen et
al. (\protect\hyperlink{ref-chen2013}{2013}) proposed approaches to fit
marginal functional models that is compatible to multi-level,
generalized outcomes. Goldsmith et al.
(\protect\hyperlink{ref-goldsmith2015}{2015}) established a model
framework that takes into account the fixed effect of time-invariant
covariates, with parameters estimated with Bayesian method in
\emph{Stan}. Gertheiss et al.
(\protect\hyperlink{ref-gertheiss2016}{2016}) identified bias introduced
by directly applying FPCA methods to generalized functions, and proposed
to address this problem using a two-stage joint estimation strategy.
Linde (\protect\hyperlink{ref-linde2019}{2009}) used an adapted Bayesian
variational algorithm for FPCA of binary and count data. In terms of
implementation, Wrobel et al. (\protect\hyperlink{ref-wrobel2019}{2019})
proposed a fast, efficient way to fit GFPCA on binary data using EM
algorithm, accompanied by the an open source R package \emph{registr}.

Existing research on dynamic prediction with functional data has focused
primarily on continuous/Gaussian outcomes, usually modelling
subject-specific random effects with FPCA
(\protect\hyperlink{ref-chiou2012}{Chiou 2012};
\protect\hyperlink{ref-goldberg2014}{Goldberg et al. 2014};
\protect\hyperlink{ref-shang2017}{Shang 2017}). Kraus
(\protect\hyperlink{ref-kraus2015}{2015}) has used this approach to
predict missing observations in partially observed function tracks, and
Delaigle and Hall (\protect\hyperlink{ref-delaigo2016}{2016}) achieved
similar goals using Markov Chains. While methods mentioned above used
only partial observations for prediction with an intercept-only model,
Leroux et al. (\protect\hyperlink{ref-leroux2016}{2018a}) proposed
Functional Concurrent Regression (FCR) framework which can incorporate
the effect of subject-specific predictors. Although methodological
development exists, flexible software for implementing these methods is
still relatively sparse with the exception of the fcr
(\protect\hyperlink{ref-fcrpkg}{\textbf{fcrpkg?}}) package accompanying
(\protect\hyperlink{ref-leroux2016}{Leroux et al. 2018a}).

In this paper, we aim to develop a fast, scalable method for dynamic
prediction of generalized functional outcomes using the principals of
local generalized linear mixed effects models combined with functional
principal components analysis of the estimated latent processes. The
proposed method is designed to handle dense, regularly measured
functional data and is extremely computationally efficient relative to
the model complexity. Section 2 explains the implementation procedure of
the proposed method. In Section 3, we illustrate the performance and
efficiency of our proposed method in a simulation study. In Section 4,
we apply this method to a real world dataset from a weight loss study.
Section 5 presents a discussion of advantages and limitation of the
proposed method.

@@

where g is an appropriate link function, \(\beta_0(t)\) is the
population mean of latent function, and \(b_i(t)\) is a subjects-level
random effect function following a zero-mean Gaussian process with
covariance operator \(\Sigma_b\). By the Karhunen-Loève theorem,
\(b_i(t)\) can be represented as the linear combination of a infinite
set of orthogonal functions:
\(b_i(t)=\sum_{k=1}^{\infty}\xi_{ik}\phi_{k}(t)\). In practice,
\(b_i(t)\) is approximated using a finite \(K\) number of basis
functions.

Estimation of \(b_i(t)\) involves estimating both
\(\{\phi_k(t): t \in \cal{T}, 1 \leq k \leq K\}\) and
\(\{\xi_{ik}: 1 \leq i \leq N, 1 \leq k \leq K\}\). For Gaussian data
this may be done by directly estimating the covariance operator
\(\Sigma_b\) using either method of moments (if the data are relatively
small) or using faster methods such as the FACE method {[}reference -
FACE{]}. However, for non-Gaussian data, direct estimation is not
possible. Instead, one may choose to estimate \(b_i(t)\) first using a
flexible spline basis (e.g.~B-splines) with dimension \(K^b\) such that
\(K^B > K\), assuming an unstructured correlation between
subject-specific spline coefficients, then orthogonalizing the resulting
estimates via eigendecomposition or singular value decomposition. This
general approach forms the basis for estimation for several GFPCA
methods {[}references - goldsmith, julia's EM algorithm{]}. However,
this approach results in rapidly escalating computational burden as both
the number of subjects, \(N\), and the density of the observed
functional predictor increases. Note that if the covariance operator
\(\Sigma_b\) is relatively complex, then \(K\) may be large and not well
approximated using simple basis functions, \(\phi_k\) (e.g.~random
intercept and slope).

Beyond the computational burden associated with fitting the model of
interest, for dynamic prediction one must then estimate the scores for a
new subject \(u\) not used in model fitting. Typically this takes the
form of the conditional expectation of the scores
\(E[\xi_{uk}|\bf{\Theta}, \bf{y}_{u}]\) where \(\bf{\Theta}\) represents
the set of parameters estimated by the model (i.e.~parameters associated
with \(\beta_0(t), \Sigma_b\)) and \(\bf{y}_u\) are the non-Gaussian
data for the new subject. Estimation of the random effects for a new
subject is non-trivial and not implemented for even very simple models
in most random effects software packages such as lme4 {[}reference{]}
and nlme {[}reference{]}. Our proposed method bypasses this problem by
using local generalized linear mixed models to estimate the covariance
operator \(\Sigma_b\) directly on the latent scale, then using the
closed form solution for the best linear unbiased predictor of Gaussian
random effects. Estimation is made fast for even very large data through
the use of the FACE methodology for FPCA on the latent estimates
obtained from local models. Because the local models are very simple
GLMMs, estimation is also fast for the local models.

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Bin the observed outcomes in to small, non-overlapping, equal length
  intervals. We hereafter index the bins by their midpoints \(s\). The
  number of observations from the same subject in bin \(s\) is \(n_s\),
  and \(t_{sj}\) refers to the jth observation point in bin \(s\).
\item
  Fit a local Generalized Mixed Model at every bin. Specifically, at bin
  \(s\), we fit \(g[E(Y_i(t_{sj}))] = \beta_0(s)+b_i(s)\),
  \(j\in\{1, ..., n_s\}\). From this series of models we can estimate
  the individual latent functions at every bin:
  \(\hat{\eta}_i(s) = \hat{\beta}_0(s)+\hat{b}_i(s)\).
\item
  Fit FPCA on the estimated latent functions:
  \(\hat{\eta}_i(s) = f_0(s)+\sum_{k=1}^K\xi_{ik}\phi_{k}(s)+\epsilon_i(s)\),
  and obtain estimates of basis functions
  \(\boldsymbol{\Phi} = \{\phi_1(s), ...,\phi_k(s)\}\), eigenvalues
  \(\hat{\lambda}_1...\hat{\lambda}_k\), mean function \(\hat{f}_0(s)\)
  and residual variance \(\hat{\sigma}^2\).
\end{enumerate}

\textbf{Ying, include the formulas for your conditional expectation here
to obtain random effects estimates }

@@

As a reference method, we compare our method to Generalized Linear Mixed
Models using Adaptive Gaussian Quadrature (GLMMadaptive). This is one of
the fastest existing method developed for dynamic prediction of repeated
generalized outcomes. Just like many mixed models, this method is very
limited in terms of flexibility if one wishes to obtain estimates in a
reasonable time frame. For example, the model used for prediction of our
simulated datset would simply be an linear model with one covariate
indicating observation time:
\(g(E(Y_i)) = \beta_0+\beta_1t+b_{i0}+b_{i1}t\). While the flexibility
of this mixed model can be increased using spline functions, the
dimension of spline basis is also restricted by computational ability,
and is unfeasible to implement under the scale of our simulated data or
the complexity of our proposed method.

@@

\begin{figure}
\centering
\includegraphics{Manuscript_edit_files/figure-latex/Figure-1.pdf}
\caption{Predicted track of four randomly selected subjects from one
simulated dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicated predicted latent
function tracks, and color indicates different observation time.}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics{Manuscript_edit_files/figure-latex/Figure_appl-1.pdf}
\caption{Predicted track of four randomly selected subjects from the
daily weigh-in dataset. The grey solid line indicates the true latent
continuous function. The dashed lines indicated predicted latent
function tracks, and color indicates different observation time.}
\end{figure}

@@

The simulation study and data application above have illustrated the
feasibility and utility of the proposed method, even for large datasets
and complex latent processes/random effect structures. can achieve
better predictive performance with much less computational burden.
Compared to competing methods, fGFPCA can accommodate extremely flexible
correlation structure between repeated measure, which is particularly
useful when outcome is generated from latent functions with highly
non-linear patterns. Due to the comparative flexibility of the fGFPCA
approach, fGFPCA has much lower ISE and higher AUC than GLMM. As
expected, the dynamic prediction accuracy of fGFPCA improves as more
data are availble for predictions. Moreover, fGFPCA also significantly
reduce time spent on both model fitting and prediction, allowing for
scaling of models and predictions which are infeasible with existing
approaches. In fact, as far as we are aware of, the proposed method is
the only feasible method that can handle the scale of the simulated
datasets with compatible flexibility.

\textbf{Andrew: I would delete this paragraph, honestly}

@@

\leavevmode\vadjust pre{\hypertarget{ref-refundpck}{}}%
Goldsmith, J., Scheipl, F., Huang, L., Wrobel, J., Di, C., Gellar, J.,
Harezlak, J., McLean, M. W., Swihart, B., Xiao, L., Crainiceanu, C., and
Reiss, P. T. (2022),
\emph{\href{https://CRAN.R-project.org/package=refund}{Refund:
Regression with functional data}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-leroux2016}{}}%
Leroux, A., Xiao, L., Crainiceanu, C., and Checkley, W. (2018a),
{``Dynamic prediction in functional concurrent regression with an
application to child growth,''} \emph{Statistics in medicine}, 37,
1376--1388.

\leavevmode\vadjust pre{\hypertarget{ref-fcrpck}{}}%
Leroux, A., Xiao, L., Crainiceanu, C., and Checkly, W. (2018b),
\emph{\href{https://CRAN.R-project.org/package=fcr}{Fcr: Functional
concurrent regression for sparse data}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-jmbook}{}}%
Rizopoulos, D. (2012), \emph{Joint models for longitudinal and
time-to-event data: With applications in r}, Chapman; Hall/CRC.
https://doi.org/\url{https://doi.org/10.1201/b12208}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wood2014}{}}%
Wood, S., and Scheipl, F. (2014), {``gamm4: Generalized additive mixed
models using mgcv and lme4,''} \emph{R package version 0.2-3}.
